\large Neural and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2023-09-15 docs/papers/sto/so.tex
FIXUP draft add two prelim figures

We sharpen the PM force with first neural networks and then symbolic
regression, taking symmetry and dimensional analysis in consideration.

@@

In cosmological $N$-body simulations, the dynamics of the particles is governed
by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
In Fourier space, the gravitational force field $- \vnabla \varphi$ reads
%
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
%
We introduce force sharpening through
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$ and $g$ are NNs, each approximating a nonlinear function of the
wavevector components and all possible relevant features $\vvartheta$ summarized
in \autoref{tab:feat} with more details discussed in \autoref{sec:features}.
Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ is the most general modification to the force
kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value for the evenness, and sort the 3 wavevector
compontents to achieve the permutation equivariance.
Being embedded in the differentiable simulation, the NNs can be trained using
gradient-based optimization methods as in usual deep learning tasks.
The parameters of the neural networks, which enter the force evaluation in
every time integration step, can be treated just the same as other cosmological
parameters.
The objective gradients are accumulated, based on which the parameters are then
optimized.
For both $f$ and $g$ neural networks, we adopt multilayer perceptron (MLP, i.e.
fully connected) structures.

@@

% transfer function & correlation coefficient for different confs
To compare two fields $\hat f$ and $f$, we introduce for each mode the square
root ratio of power
%
\begin{equation}
T(\vk) \triangleq
\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk)|^2}},
\label{T}
\end{equation}
%
which is also referred to as the transfer function, and the cross correlation
coefficient
%
\begin{equation}
r(\vk) \triangleq
\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
     {\sqrt{|\hat f(\vk)|^2 |f(\vk)|^2}}.
\label{r}
\end{equation}
%
The two fields are identical when both $T$ and $r$ are unity.

@@

We empoly a learning rate scheduler that reduces the learning rate on plateau
where the loss function stops decreasing.

>>>>

We optimize the spatial resolution of cosmological particle-mesh (PM)
simulations by sharpening the PM force with first neural networks and then
symbolic regression, taking symmetry and dimensional analysis in consideration.

@@

In cosmological $N$-body simulations, the dynamics of the particles is governed
by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
In Fourier space, the gravitational force field $- \vnabla \varphi$ reads
%
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
%
where we introduce force sharpening for spatial optimization through
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$ and $g$ are NNs, each approximating a nonlinear function of the
wavevector components and all possible relevant features $\vvartheta$ summarized
in \autoref{tab:feat} with more details discussed in \autoref{sec:features}.
Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ is the most general modification to the force
kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value for the evenness, and sort the 3 wavevector
compontents to achieve the permutation equivariance.
Being embedded in the differentiable simulation, the NNs can be trained using
gradient-based optimization methods as in usual deep learning tasks.
The parameters of the neural networks, which enter the force evaluation in
every time integration step, can be treated just the same as other cosmological
parameters.
The objective gradients are accumulated, based on which the parameters are then
optimized.
For both $f$ and $g$ neural networks, we adopt multilayer perceptron (MLP, i.e.
fully connected) structures.

@@

\begin{figure*}
  \centering
  \includegraphics[width=.8\columnwidth]{sobol0snap120epoch300.pdf}
  \caption{The transfer function $T$ and correlation coefficient $r$ of the
  \GADGET\ and \pmwd\ simulated density fields, with and without force
  sharpening.
  Sobol 0 snap 120}
  \label{fig:tfcc}
\end{figure*}

% transfer function & correlation coefficient
To compare two fields $\hat f$ and $f$, we introduce for each mode the square
root ratio of power
%
\begin{equation}
T(\vk) \triangleq
\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk)|^2}},
\label{T}
\end{equation}
%
which is also referred to as the transfer function, and the cross correlation
coefficient
%
\begin{equation}
r(\vk) \triangleq
\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
     {\sqrt{|\hat f(\vk)|^2 |f(\vk)|^2}}.
\label{r}
\end{equation}
%
The two fields are identical when both $T$ and $r$ are unity.
With force sharpening,

\begin{figure*}
  \centering
  \includegraphics[width=.8\columnwidth]{denslab_0_120_e300.pdf}
  \caption{Projected dark matter density fields of \GADGET, \pmwd\ without and
  with force sharpening.
  Sobol 0 snap 120}
  \label{fig:denslab}
\end{figure*}

@@

We empoly a learning rate scheduler that reduces the learning rate on plateau
where the loss function stops decreasing.
The initial learning rate is set at [...] with a reduction factor of [...] after
[...] epochs of no decreasing on loss.
